MSD panels 1, 2 and 3 Explant Supernatant

### 2016-09-20 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
Changed the below det values to 0.01, subtracted mock from the virus data

---
title: "MSD panels 1, 2 and 3 HVE Sup"
author: "Claire Levy"
date: "August 22, 2016"
output: github_document
---

@@

```{r standards, echo=FALSE, message = FALSE, warning = FALSE}

@@

ggplot(HVE_stds_all_panels_percent_recov,aes( x = Sample, y = X..Recovery)) +
  geom_point(aes(shape = Detection.Range ),
             size = 2, alpha = 0.5)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, 
                                "In Detection Range" = 16, 
                                "Below Detection Range" = 0))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(0, 265))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = rel(0.8)))

@@

```{r checking detection range,fig.height = 6, fig.width = 10, echo=FALSE, message = FALSE, warning = FALSE}

@@

ggplot(HVE_all_panels_noStds_NumInEach, aes(x = Assay, y=Number_of_Samples))+
         geom_point(aes(color = Detection.Range),size = 2)+
  theme(axis.text.x = element_text(angle=25, size = 8))+
  labs(y = "Number of Samples", x = "Analyte")+
  facet_wrap(Sample_Dilution~Panel, scales = "free")+
  ggtitle("Number of Samples in each Detection Range per Analyte")
```

@@

## All Samples Estimated Stock Concentrations

@@

```{r  all sample concentrations, echo=FALSE, message = FALSE, warning = FALSE}

@@

ggplot(HVE_all_panels_noStds_noBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
         geom_point(aes(color = Assay),position = position_swarm (0.5),size = 2)+
  scale_y_log10()+
  labs(y = "Estimated Stock Concentration pg/ml")+
  facet_wrap(~Virus)

@@

## Plots of analyte concentrations
Since the 1:2 diluted samples were more likely to be in the detection range, I am only showing data for those.

@@

```{r  individual analyte plots,fig.width = 8, fig.height= 13 , echo=FALSE, message = FALSE, warning = FALSE}

ggplot(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
   geom_point(aes(color = Donor), alpha = 0.75, size = 3)+
  scale_y_log10()+
   labs( y = "Estimated Stock Concentration pg/ml")+
   facet_grid(Assay~Virus, scales = "free_x")+
   theme(axis.text.x = element_text(size = 11))

>>>>

---
title: "MSD panels 1, 2 and 3 HVE Supernatant"
output: pdf_document
---

@@

```{r standards,fig.height = 7, fig.width = 12, echo=FALSE, message = FALSE, warning = FALSE}

@@

ggplot(HVE_stds_all_panels_percent_recov,aes( x = Sample, y = X..Recovery)) +
  geom_point(aes(shape = Detection.Range ),
             size = 2, alpha = 0.5)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, 
                                "In Detection Range" = 16, 
                                "Below Detection Range" = 0))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(0, 265))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = 9))+
  ggtitle("Percent Recovery of Standards")

@@

```{r checking detection range, fig.height=8, fig.width =12, echo=FALSE, message = FALSE, warning = FALSE}

@@

ggplot(HVE_all_panels_noStds_NumInEach, aes(x = Assay, y=Number_of_Samples))+
         geom_point(aes(color = Detection.Range),size = 2)+
  theme(axis.text.x = element_text(angle=25, size = 11))+
  labs(y = "Number of Samples", x = "Analyte")+
  facet_wrap(Sample_Dilution~Panel, scales = "free")+
  ggtitle("Number of Samples in each Detection Range per Analyte")
```

@@

## Estimated Sample Concentration Distribution

@@

```{r  all sample concentrations,fig.height = 6, fig.with = 10,  echo=FALSE, message = FALSE, warning = FALSE}

@@

ggplot(HVE_all_panels_noStds_noBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
         geom_point(aes(color = Assay),position = position_swarm (0.5),size = 2)+
  scale_y_log10()+
  labs(y = "Estimated Stock Concentration pg/ml")+
  facet_wrap(~Virus)+
  ggtitle("Estimated Sample Concentration Distribution by Treatment and Timepoint")

@@

## Plots of Analyte Concentrations by Donor, Timepoint and Condition

Since the 1:2 diluted samples were more likely to be in the detection range, I am only showing data for those.

@@

```{r  individual analyte plots, echo=FALSE,fig.width=8, fig.height = 10, message = FALSE, warning = FALSE}

ggplot(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
   geom_point(aes(color = Donor), alpha = 0.75, size = 3)+
  scale_y_log10()+
   labs( y = "Estimated Stock Concentration pg/ml")+
   facet_grid(Assay~Virus, scales = "free_x")+
   theme(axis.text.x = element_text(size = 11))+
  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")

### 2016-09-20 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Changed the below det values to 0.01, subtracted mock from the virus data

---
title: "MSD panels 1, 2 and 3 Explant Sup"
author: "Claire Levy"
date: "August 12, 2016"
output: github_document
---

@@

#remember that this data table contains data from the 17Aug16_HVE_sup_panel_1_plate_1. I am only going to use the Standards data from that plate and not the standards from either of the plates for panel 1 run 10Aug16 since we had what appears to be a pipetting error (2nd replicate is more concentration for all wells).

@@

ggplot(explant_stds_all_panels_percent_recov,aes( x = Sample, y = X..Recovery)) +
  geom_point(aes(shape = Detection.Range ),
             size = 2, alpha = 0.5)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, 
                                "In Detection Range" = 16, 
                                "Below Detection Range" = 0))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(0, 250))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = rel(0.8)))

@@

#Add a column for Sample dilution (all samples were diluted 1:2)

donorTimeVirusDil <- donorTimeVirus %>%
  mutate(Sample_Dilution = "1:2")

@@

#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep (sample 324-T2-V1 where I ran out of sample before I could do the 2nd rep)

@@

```{r checking detection range,fig.height = 5, fig.width = 10, echo=FALSE, message = FALSE, warning = FALSE}

@@

@@

#I'm going to make a column for estimated stock concentration by multiplying the avgConc by 2 (since I did a 1:2 dilution on the samples) so I can better compare with the HVE sup concentrations.

explant_all_panels_noStds <- explant_all_panels_noStds%>%
  mutate(Estimated_Stock_Conc =  Calc..Concentration * 2)

@@

```{r exclude belows, echo=FALSE, message = FALSE, warning = FALSE}

#I am also excluding samples that are "Below Detection Range" because this means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable."

explant_all_panels_noStds_noBelow <- explant_all_panels_noStds %>%
  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")
```

## All Samples Estimated Stock Concentrations

@@

```{r  all sample concentrations, echo=FALSE, message = FALSE, warning = FALSE}

#Plot of all samples and their concentrations

ggplot(explant_all_panels_noStds_noBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
         geom_point(aes(color = Assay),position = position_swarm (0.5),size = 2)+
  scale_y_log10()+
  labs(y = "Estimated Stock Concentration pg/ml")+
  facet_wrap(~Virus)

@@

#averaging the replicates

#some of the samples had one rep that was below detection, so I am excluding those so I will be able to use the remaining rep as the "average" for that sample.

avgConc_explant_all_panels_noStds_noBelow <- explant_all_panels_noStds_noBelow %>%
group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
summarize(avgEstConc = mean(Estimated_Stock_Conc))%>%
  ungroup()

#Looking at the number of samples in each sample "set"(Same Donor and Timepoint). Some are missing V1 or V2. or both!! I don't want the ones with just one sample in the set.

numberInSets<-avgConc_explant_all_panels_noStds_noBelow %>%
         group_by(Donor,Timepoint,Assay)%>%
  select(Sample,Donor, Timepoint,Assay)%>%
  summarize(numInSet = n())

#now I'll merge the avgConc df with this numberInSets info created above so I can then filter out samples that have <2 samples in the set

avgConc_explant_all_panels_noStds_noBelow <- merge(avgConc_explant_all_panels_noStds_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))

#now exclude the samples where there not a "full set", i.e. 1 sample only in the set

avgConc_explant_all_panels_noStds_noBelow_fullSet <- avgConc_explant_all_panels_noStds_noBelow %>%
  filter(numInSet != 1)

@@

## Plots of analyte concentrations

We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock suparnatent concentration.

@@

* Concentrations below the fit curve range. Values cannot be extrapolated when the data was *below* the fit curve range. However,the MSD software can extrapolate values that are *above* the fit curve range if the curve is linear at the top.

* Samples for which there was only data for a Mock sample and neither of the virus conditions.

@@

ggplot(avgConc_explant_all_panels_noStds_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
  scale_y_log10()+
   labs( y = "Estimated Stock Concentration pg/ml")+
   facet_grid(Assay~Virus, scales = "free_x")+
   theme(axis.text.x = element_text(size = 11))

################## another way of making the plots #####

# #make a list of the analytes
# 
# #I'm using a unique vector here instead of the levels of Assay because I don't want to include IFNa2a OR Il-7. There weren't any data points in the detection range for those analytes (so I don't want to try to make a plot of it)
# 
# 
# analyteList<-as.list(unique(avgConc_explant_all_panels_noStds_noBelow$Assay))
# 
# 
# 
# 
# #write a function to make a plot for each analyte
# 
# plotAnalytes <- function (Analyte){
#    avgConc_explant_all_panels_noStds_noBelow_fullSet %>%
#      filter(Assay == Analyte)%>%
#      
#      ggplot( aes(x = Timepoint, y = Estimated_Stock_Conc))+
#    
#    geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
#   geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
#     stat_summary(fun.y = mean, geom = "point")+
#     stat_summary(fun.y = mean, geom = "line",aes(group = 1))+
#    
#    scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+
#   
#    scale_y_log10()+
#    
#    labs( y = "Concentration pg/ml")+
#    
#   facet_wrap(~Virus, scales = "free_x")+
#    
#    theme(axis.text.x = element_text(size = 11))+
#      ggtitle(Analyte)
# }
# 
# 
# 
# lapply(analyteList, FUN = plotAnalytes)

>>>>

---
title: "MSD panels 1, 2 and 3 Explant Supernatant"
output: pdf_document
---

@@

#remember that this data table contains data from the 17Aug16_HVE_sup_panel_1_plate_1. I am only going to use the Standards data from that plate (not the data) and not the standards from either of the plates for panel 1 run 10Aug16 since we had what appears to be a pipetting error (2nd replicate is more concentrated for all wells).

@@

ggplot(explant_stds_all_panels_percent_recov,aes( x = Sample, y = X..Recovery)) +
  geom_point(aes(shape = Detection.Range ),
             size = 2, alpha = 0.5)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, 
                                "In Detection Range" = 16, 
                                "Below Detection Range" = 0))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(0, 250))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = 9))+
  ggtitle("Percent Recovery of Standards")

@@

#Add a column for Sample dilution (all samples were diluted 1:5)

donorTimeVirusDil <- donorTimeVirus %>%
  mutate(Sample_Dilution = "1:5")

@@

#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep (sample 324-T2-V1 where I ran out of sample before I could do the 2nd rep for panel 1. there wasnt enough of this sample to use any for panel 2 :( )

@@

```{r checking detection range,fig.height = 4, fig.width = 12, echo=FALSE, message = FALSE, warning = FALSE}

@@

#Were there any samples where none of the assays seemed to have worked 
#i.e. All 40 were below?
explant_all_panels_noStds_NumInEach_bySample <- explant_all_panels_noStds%>%
 group_by(Sample, Detection.Range)%>%
  summarise(Number_of_Reps = n())%>%
  filter(str_detect(Detection.Range, "Below"))%>%
  group_by(Sample)%>%
  summarise(Number_of_Belows = sum(Number_of_Reps))
#None of these numbers are 40, so it is ok.

@@

#I'm going to make a column for estimated stock concentration by multiplying the avgConc by 5 (since I did a 1:5 dilution on the samples) so I can better compare with the HVE sup concentrations.

explant_all_panels_noStds <- explant_all_panels_noStds%>%
  mutate(Estimated_Stock_Conc =  Calc..Concentration * 5)

@@

```{r adjust belows, echo=FALSE, message = FALSE, warning = FALSE}

#Samples that are "Below Detection Range" means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable."

#HOWEVER, we still want to show the fact that those samples had really low concentrations. So, I am going to set all those samples to 0.001 so they show up on plots.

#which ones were those again?

explant_belows <- explant_all_panels_noStds %>%
  filter(str_detect(Detection.Range, "Below"))%>%
  group_by(Donor, Timepoint, Assay)%>%
  summarize(Number_of_samples = n())

#I'm going to set all of the "belows" to 0.001

explant_all_panels_noStds_AdjBelow <- explant_all_panels_noStds %>%
mutate(Estimated_Stock_Conc = ifelse(str_detect(Detection.Range,"Below"),0.01, Estimated_Stock_Conc))

##Estimated Sample Concentration Distribution

@@

```{r  all sample concentrations,fig.height = 6, fig.with = 10, echo=FALSE, message = FALSE, warning = FALSE, results = 'asis'}

#Plot of all samples and their concentrations
#Note that this includes the samples that manually set to 0.01

ggplot(explant_all_panels_noStds_AdjBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
         geom_point(aes(color = Assay),position = position_swarm (0.5),size = 2)+
  scale_y_log10()+
  labs(y = "Estimated Stock Concentration pg/ml")+
  facet_wrap(~Virus)+
  ggtitle("Estimated Sample Concentration Distribution by Treatment and Timepoint")

cat("\n\n\\pagebreak\n")

@@

#First change the "Below..." entries to just say "Below", since I don't care which "Below..." it was for my purposes.

explant_all_panels_noStds_AdjBelow <- explant_all_panels_noStds_AdjBelow %>%
  mutate(Detection.Range = ifelse(str_detect(Detection.Range, "Below"), "Below", as.character(Detection.Range)))

#which samples have discordant replicates in terms of det.range?
repsInRange<-explant_all_panels_noStds_AdjBelow%>%
  group_by(Sample, Assay, Detection.Range)%>%
  summarize(Number_Reps = n())

#merge the repsInRange data with the explant_all_panels_noStds_AdjBelow data by the Sample and Assay

#Result is a df where the samples with dischordant replicates in terms of Detection.Range. The samples with "1" in the Number_of_Reps colimn are the ones where I want to use the data from the rep that was in/above range for the avg, NOT the average between the in-range rep and the below-range rep. Recall that I assigned all below range/curve reps a value of 0.01. If both reps were below, ok to use calculate the average of them.

# select fewer cols for readability.

mergeRepsInRange <- explant_all_panels_noStds_AdjBelow %>%
  merge(.,repsInRange, by = c("Sample", "Assay", "Detection.Range"))%>%
  select(Sample,Timepoint,Donor,Virus,Assay,Detection.Range, Estimated_Stock_Conc, Number_Reps, Plate.Name)

#HOWEVER: I know that for sample 324-T2-V1, I only had one replicate for the assays on panel 2, so for just this sample, I don't want to eliminate a rep that meets this exclusion criteria (IL-12p70) because I don't actually have a good rep to use as the average. I'll add it back in after filtering.

T2V1324<-mergeRepsInRange %>%
  filter(Sample == "324-T2-V1", Assay =="IL-12p70")

#The  rows of data that I want will fit the following criteria:

#Detection.Range = Below and Number_reps = 2 (both reps are below)
#Detection.Range != Below and Number_reps = 2 (both reps are good)
#Detection.Range != Below and Number_reps = 1 (The in/above range rep from a discordant set)

#I am doing this by filtering for the things I want to keep, then combining those dfs, rather than writing one expression to say what to exclude. (mostly because I couldnt figure it out...)

bothBad <- mergeRepsInRange %>%
  filter(Detection.Range =="Below"& Number_Reps ==2)

discordGood<- mergeRepsInRange %>%
  filter(Detection.Range != "Below"& Number_Reps ==1)

bothGood <- mergeRepsInRange %>%
  filter(Detection.Range != "Below"& Number_Reps == 2)

#combine all the reps

OKtoAvg <- rbind(bothBad,discordGood, bothGood,T2V1324)

#averaging the replicates
avgConc_explant_all_panels_noStds_AdjBelow <- OKtoAvg %>%
group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
summarize(avgEstConc = mean(Estimated_Stock_Conc))%>%
  ungroup()

@@

```{r  samples in sets, fig.width = 8, fig.height= 13, echo=FALSE}

#Looking at the number of samples in each sample "set"(Same Donor and Timepoint).

numberInSets<-avgConc_explant_all_panels_noStds_AdjBelow %>%
         group_by(Donor,Timepoint,Assay)%>%
  select(Sample,Donor, Timepoint,Assay)%>%
  summarize(numInSet = n())

#now I'll merge the avgConc df with this numberInSets info so I know which conditions are in each set.
avgConc_explant_all_panels_noStds_AdjBelow <- merge(avgConc_explant_all_panels_noStds_AdjBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))

# For 324-T2, there are ONLY mock samples
# There are no samples with just V1 and/or V2 and no mock

```{r  subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}

#I want to subtract out the mock data from the corresponding v1 and V2 data.

#extract out Mock samples

mock <- avgConc_explant_all_panels_noStds_AdjBelow %>%
  filter(Virus == "M")

# remove mock from the rest of the data

no_mock_avgConc_explant_all_panels_noStds_AdjBelow <-
  avgConc_explant_all_panels_noStds_AdjBelow %>%
  filter(Virus !="M")

mergeMock<-merge(no_mock_avgConc_explant_all_panels_noStds_AdjBelow, mock, by = c("Donor", "Assay", "Timepoint"), all.x = TRUE, all.y = TRUE)

#check for NA's (places where there was a mock sample but no corresponding V1 or V2)
whichNAs<-mergeMock[which(is.na(mergeMock$Sample.x)),]
#This is just donor 324 Il1ra, mig and rantes, which we already know has just mock.
# Remove the NAs for tidyness

mergeMock<-mergeMock%>%
  na.omit()%>%
  select(Donor, Assay, Timepoint, Sample.x, Virus.x, avgEstConc.x, numInSet.x, avgEstConc.y)

#Fix up the colnames

names(mergeMock)[4:8]<- c("Sample","Virus","avgEstConcVirus","numInSet","avgEstConcMock")

#Do the subtraction

mergeMock<- mergeMock %>%
  mutate(minusMockEstConc = avgEstConcVirus-avgEstConcMock)

#When the experimental and mock samples were both "below", we get zero in the minusMockEstConc column, and that won't show up on a log scale. So I'm setting all zeros in the minusMockEstConc column to 0.01

mergeMock <- mergeMock %>%
  mutate(minusMockEstConc = ifelse(minusMockEstConc <= 0, 0.01, minusMockEstConc))

anyNA(mergeMock)
```

## Plots of Analyte Concentrations by Donor, Timepoint and Condition
We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock suparnatent concentration.

@@

* Samples for which there was only data for a Mock sample and neither of the virus conditions (Donor 324, IL-1RA, MIG, RANTES)

@@

ggplot(mergeMock, aes(x = Timepoint, y = minusMockEstConc))+
   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
   scale_y_log10()+
   labs( y = "Estimated Stock Concentration pg/ml")+
   facet_grid(Assay~Virus, scales = "free_x")+
   theme(axis.text.x = element_text(size = 11))+
  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
